---
title: "BGC_modelR"
format: html
editor: visual
---

Translating this Python script to R requires adapting the classes, configuration handling, and geospatial operations using R’s packages such as `data.table`, `sf`, and `terra`. Here is an R equivalent for key parts of the script, emphasizing efficient data manipulation and spatial processing.
This R script provides an outline for key processes from the Python script, focusing on R functions in sf and terra for spatial processing and raster manipulation, and data.table for fast data handling. Additional model-specific details (such as classifications and configurations) can be implemented in the initialize_bec function and other helper functions to match the specific needs of the BEC model translation.

### 1. Load Necessary Libraries

```{r}
library(data.table)
library(sf)
library(terra)
library(lubridate)
library(config)
library(jsonlite)
```

2\. Configurations and Custom Error Handling

```{r}
# Load configurations using a JSON or YAML configuration file.
config <- config::get(file = "config.yml")

# Define custom error functions for config validation
config_error <- function(message) stop(paste("Configuration Error:", message))
config_value_error <- function(message) stop(paste("Configuration Value Error:", message))

# Validate the configuration
validate_config <- function(config) {
  required_keys <- c("rulepolys_file", "elevation", "cell_size_metres", "temp_folder")
  
  for (key in required_keys) {
    if (is.null(config[[key]]) || !file.exists(config[[key]])) {
      config_value_error(paste(key, "does not exist or is not configured."))
    }
  }
  
  if (config$cell_size_metres < 25 || config$cell_size_metres > 100 || config$cell_size_metres %% 5 != 0) {
    config_value_error("cell_size_metres must be a multiple of 5 between 25 and 100.")
  }
}
validate_config(config)

```
3. Load Spatial Data
```{r}
# Load rule polygons and elevation data as spatial objects
rulepolys <- st_read(config$rulepolys_file)
elevation <- fread(config$elevation)

# Define workspace and output directories
wksp <- config$temp_folder
if (!dir.exists(wksp)) dir.create(wksp, recursive = TRUE)

# Load DEM (Digital Elevation Model) data
dem <- if (file.exists(config$dem)) {
  rast(config$dem)
} else {
  rast(system.file("external/test.grd", package="terra")) # Placeholder if DEM does not exist
}

```

4. Aspect and Slope Calculations
```{r}
# Calculate slope and aspect using terrain functions
slope <- terrain(dem, opt = "slope", unit = "degrees")
aspect <- terrain(dem, opt = "aspect", unit = "degrees")

```

5. Rasterize Rule Polygons

```{r}
# Rasterize rule polygons to align with DEM resolution
rule_raster <- rasterize(rulepolys, dem, field = "polygon_number", touches = TRUE)
rule_expanded <- focal(rule_raster, w = matrix(1, 3, 3), fun = "mean")  # expand by a specified factor

```

6. Initial Ecosystem Classification Raster
```{r}
# Initialize BEC classification raster based on elevation, aspect, rule polygons
initialize_bec <- function(elevation, rule_raster, aspect, slope) {
  bec_raster <- rast(rule_raster)
  values(bec_raster) <- NA
  
  for (row in 1:nrow(elevation)) {
    # Define aspect zone and elevation ranges
    cool_range <- c(elevation[row, cool_low], elevation[row, cool_high])
    warm_range <- c(elevation[row, warm_low], elevation[row, warm_high])
    
    # Apply elevation and aspect-based classifications
    bec_raster <- classify(bec_raster, cool_range, elevation[row, becvalue])
  }
  
  return(bec_raster)
}

bec_raster <- initialize_bec(elevation, rule_raster, aspect, slope)

```

7. Post-Filtering: Majority Filter and Noise Reduction
```{r}
# Apply majority filter to smooth raster classifications
apply_majority_filter <- function(bec_raster, slope, config) {
  low_slope_filter <- matrix(1, config$majority_filter_size_slope_low_metres, config$majority_filter_size_slope_low_metres)
  steep_slope_filter <- matrix(1, config$majority_filter_size_slope_steep_metres, config$majority_filter_size_slope_steep_metres)
  
  filtered_raster <- focal(bec_raster, w = ifelse(slope < config$steep_slope_threshold, low_slope_filter, steep_slope_filter), fun = "modal")
  return(filtered_raster)
}

majority_filtered <- apply_majority_filter(bec_raster, slope, config)

```

8. Noise Filtering and Hole Filling

```{r}
# Use connected component labeling for noise removal in high-elevation zones
noise_filtered <- classify(majority_filtered, c(0, config$noise_removal_threshold), NA, others = FALSE)

# Fill small holes in zones to smooth classifications
noise_filled <- focal(noise_filtered, w = matrix(1, 3, 3), fun = "modal", na.rm = TRUE)

```
9. Save Output to GeoPackage or Shapefile
```{r}
# Convert raster to polygons and save to output file
bec_polygons <- as.polygons(noise_filled, dissolve = TRUE)
bec_polygons <- st_as_sf(bec_polygons)

# Add area and labels
bec_polygons$area_ha <- st_area(bec_polygons) / 10000  # Convert to hectares
bec_polygons$label <- "Classification"  # Placeholder for actual labels

# Write to output file
output_file <- config$out_file
st_write(bec_polygons, output_file, driver = ifelse(grepl(".gpkg$", output_file), "GPKG", "ESRI Shapefile"))

```

